---
title: "Model Definitions and Analysis"
author: "Vassar CogSci Lab Team"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
bibliography: ../references.bib
output:
  rmarkdown::html_vignette:
    css: ../stile.css
    toc: false
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Model Definitions and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  error = TRUE,
  comment = "#>"
)
```

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

<div id="TOC">
<ul>
<li><a href="#exp">The Experiment (Data Collection)</a></li>
<li><a href="#def">Model Definitions</a>
  <ul>
  <li><a href="#def-no">No Learning</a></li>
  <li><a href="#def-step">Step Learning</a></li>
  <li><a href="#def-sym">Symmetric Logistic Learning</a></li>
  <li><a href="#def-asym">Asymmetric Logistic Learning</a></li>
  </ul></li>
<li><a href="#rec">Model Recovery</a>
  <ul>
  <li><a href="#rec-nolearn">No Evidence of Learning</a></li>
  <li><a href="#rec-learn">Strong Evidence of Learning</a></li>
  <li><a href="#rec-small">Consistent but Small Decrease in Response Times</a></li>
  <li><a href="#rec-ret">Sporadic Return to Baseline Response Time</a></li>
  <li><a href="#rec-inc">Increased Response Times at the End of the Experiment</a></li>
  <li><a href="#rec-other">Other Considerations</a></li>
  </ul></li>
<li><a href="#comp">Formal Model Comparison with PSIS-LOO</a>
  <ul>
  <li><a href="#comp-fit">Fit the Models to Data</a></li>
  <li><a href="#comp-pre">Preliminary Results of LOO</a></li>
  <li><a href="#comp-more">More Results</a></li>
  </ul></li>
<li><a href="#references">References</a></li>
</ul>
</div>






The Vassar CogSci Lab is modelling some pretty cool stuff.
<br><br>

# Some equation with a paragraph

This is just a sample so I know how to do it later.

\begin{equation} \label{eq:kl-Nav}
	k_\ell^{\text{Nav}} \left( t', \epsilon \right) = \left\lceil \max \left\{ \sqrt{\frac{-2 \log(\pi t' \epsilon)}{\pi^2 t'}}, \frac{1}{\pi \sqrt{t'}} \right\} \right\rceil.
\tag{L.1}
\end{equation}

The small-time approximations, Equations <span class="math inline eqref">\(\eqref{eq:kl-Nav}\)</span> and





# The Experiment (Data Collection) {#exp}
<hr class="sec1">

This is my understanding of how the data were collected in "Experiment 1" from Josh's thesis.

Participants were presented with a sequence of letters moving across window on a computer screen and were supposed to type the letter on the computer's keyboard. The letters appeared animated and moved across the window from right to left, and they were only visible on the screen for two seconds. The response time for each typed letter was bounded below by zero seconds because the timing only started from the moment when the letter first started appearing on the right side of the window, well before the full letter was easily visible. The response times were also bounded above by two seconds because the letters were only visible on the screen for two seconds. In these sequences of letters, there were three different types of three-letter patterns that the participant could encounter. The patterns were sorted by difficulty and classified as one of: easy, three-letter words in the English language; medium, three-letter words in the English language but spelled backwards, or hard, a random permutation of a three-letter English word but was not itself a word in the English language. There was only one intentional pattern in the sequence of letters for each participant, and this pattern would randomly appear approximately seventy times in the sequence throughout the experiment. For more details on the experiment, see @de2016dynamic.

Of particular interest to us are the sequences of response times for the first letter in the pattern and the third letter in the pattern. The idea is that since the first letter in the pattern is randomly placed in the sequence, the participant will exhibit a response time for the first letter that is consistent with their baseline response time for the rest of the random letters in the sequence. If the participant learns the pattern, we expect their response times for the first letter to remain consistent with their response times for the random letters in the sequence; however, we expect their response times for the subsequent letters (in particular the third letter) to decrease as they are able to anticipate this letter and begin to move their finger to the appropriate key before the letter appears on the screen. In the case that te participant does not learn the pattern among the sequence of letters, we expect their response times for every letter in the sequence to be consistent with their baseline response time for the random letters in the sequence. Our main objective with this model is to capture the *possible* decrease in response time for the third letter in the pattern, as this change indicates anticipatory behavior and recognition of the pattern.

Just a quick question about experiment design (because now I'm starting to think about it)- if there were experiment that follows the same setup to this one, but after showing the pattern many times (ideally after the participant learned the pattern, wonder how we could tell if they had!) the third letter in the pattern was changed to something random? I don't know if there's any point in this, but if the response time for the changed letter is higher than that for the baseline random letter would that also demonstrate that the individual had learned the pattern? I'm kinda thinking in the context of a teacher assigning math problems for homework, so maybe it's too different because the "response time" for answering a math problem is likely greater than two seconds. But if you could measure how long it takes a student to do a particular type of problem and give the student a sequence of problems that are all similar, wouldn't we expect to see a decrease in "response time" when the student figures out how to do that type of problem? Then to test if they actually learned the best strategy for solving that type of problem (instead of recognizing the pattern of what to do), the student could then be presented with a slightly different problem that requires a different strategy (perhaps something they haven't learned yet). If the student just recognized the pattern of the previous problems instead of actually learning the material, then I would expect their response time to be consistent with their decreased learned response times (and their answer to be incorrect). On the other hand, if the student has worked out the strategy for the previous type of problem and understands the material, then I would think that their response time should be much greater because they are trying to figure out why their previous strategy doesn't work. Maybe this would be totally unfeasible to ask kids to do a ton of math problems even though I'd probably enjoy it, and maybe it doesn't work for a number of other reasons, but I thought I'd ask about it anyway.





# The Models {#def}
<hr class="sec1">

The starting point of these models is to define how the response times are distributed. As discussed by @dablander2019bayesian, we treat each response time from each subject to be Log-Normally distributed with parameters $\mu$ and $\sigma$; we'll write it as $rt \sim \mathcal{LN}(\mu, \sigma)$. The Log-Normal distribution has some useful properties for our application and we'll take a quick look at some of them, but for extended information on the Log-Normal distribution see [its Wikipedia page](https://en.wikipedia.org/wiki/Log-normal_distribution). Primarily, the Log-Normal distribution has no support for negative response times, yet it resembles something of a normal distribution; these properties are consistent with the experimental data.

The mean, mode, and variance of the distribution are all slightly messy expressions involving both $\mu$ and $\sigma$, but the median of the distribution is a very clean $\exp(\mu)$. In each of the expressions for these basic properties, the parameters $\mu$ and $\sigma^2$ appear exponentiated. To exploit this structure and make our parameterization more convenient, we instead use the transformed parameters $\hat{\mu} = \log(\mu)$ and $\hat{\sigma^2} = \log(\sigma^2)$. This transformation allows us to easily eliminate the exponentiation in the properties, and in particular it allows us to work directly with the median:
\begin{equation} \label{eqn:log-hat}
  \begin{aligned}
    \text{med}\big(rt \sim \mathcal{LN}(\hat{\mu}, \hat{\sigma^2})\big) &= \exp\left[ \hat{\mu} \right] = \mu,\\[3ex]
    \text{mean}\big(rt \sim \mathcal{LN}(\hat{\mu}, \hat{\sigma^2})\big) &= \exp\left[ \hat{\mu} + \frac{1}{2} \hat{\sigma^2} \right] = \mu \sigma,\\
    \text{mode}\big(rt \sim \mathcal{LN}(\hat{\mu}, \hat{\sigma^2})\big) &= \exp\left[ \hat{\mu} - \hat{\sigma^2} \right] = \frac{\mu}{\sigma^2},\\
    \text{var}\big(rt \sim \mathcal{LN}(\hat{\mu}, \hat{\sigma^2})\big) &= \left( \exp\left[ \hat{\sigma^2} \right] - 1 \right) \exp\left[ 2 \hat{\mu} + \hat{\sigma^2} \right] = \left( \sigma^2 - 1 \right) \mu^2 \sigma^2.
  \end{aligned}
  \nonumber
\end{equation}

Since we will use the transformed parameter $\hat{\mu}$, our characterization of $\mu$ will equal the median of the distribution. We'll build the model around this fact; when we define a functional form for $\mu$, we're actually defining the behavior of the median of the response time distribution. The main goal of the model is to capture a *possible* consistent decrease in response time after a specific trial, or in other words to identify a decrease in the median response time. This is why we're going to work with $\hat{\mu}$ &ndash; so that we can directly change the median of the response time distribution.

A (potentially beneficial) side effect of working with $\hat{\mu}$ is that a decrease in $\hat{\mu}$ causes not only a decrease in the median response time, but it also causes a decrease in the variance of the response times (see the bottom equation in the above block). Josh and I had discussed this feature of the data a long time ago, but it seems to be logical that the response times are less variant once the participant learns the pattern. I haven't placed much emphasis on $\sigma^2$ because it mainly controls the variance of the response time distribution, so it just gets fit with the data.

Each individual has response time data for both non-learned letters (denoted as $rt_n$) and learned letters (denoted $rt_\ell$). The response times of the non-learned letters should be consistent with the baseline response time for a random letter as the start of the pattern is randomly placed in the sequence. In contrast, the response times of the learned letters may be consistently shorter than the baseline as the participant may be able to anticipate the letters that appear later in the pattern. Since we could potentially have two distinct clusters of response times, we will fit these two clusters of response times separately; this yields fitting two distinct $\hat{\mu}$ (i.e., $\hat{\mu_n}$ and $\hat{\mu_\ell}$), but we will use the same $\hat{\sigma^2}$ for each cluster. For each individual, the response time data will be modeled as
\begin{equation} \label{eqn:models}
  \begin{aligned}
    rt_n &\sim \mathcal{LN}(\hat{\mu_n}, \hat{\sigma^2}),\\[2ex]
    rt_\ell &\sim \mathcal{LN}(\hat{\mu_\ell}, \hat{\sigma^2}).\\[2ex]
  \end{aligned}
  \tag{M.0}
\end{equation}

The following subsections will define functional forms for $\hat{\mu_n}$ and $\hat{\mu_\ell}$ that enable the model to capture the evidence of learning (or lack of evidence) shown by each participant in the study. We will start with the simplest model and gradually add more features. Since the models will successively build on each other, the same parameters will arise in multiple models; although repetitive, we include the definitions of all parameters for completeness, but note that the old parameters will never be redefined. As we introduce each model, we will load the pre-compiled Stan model into the `R` workspace.

```{r preamble, eval=TRUE}
# load rstan and set options
library("rstan")
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=corei7 -mtune=corei7')

# set variable for filepaths depending on the operating system
if (.Platform$OS.type == "windows") { os <- "windows" } else { os <- "linux" }
```

Even though we will not read and compile the models from scratch here, the `R` code to do so is available in the raw RMarkdown file.

```{r read-stan, eval=FALSE, include=FALSE}
# no learning
no_learning_model_stanc <- stanc(
  file = "stan-models/no_learning_model.stan",
  model_name = "no_learning_model")
no_learning_model <- stan_model(
  stanc_ret = no_learning_model_stanc)
save(no_learning_model,
  file = paste0("stan-models/compiled/", os,
    "/no_learning_model.Rds"),
  compress = "xz", compression_level = 9)

# step learning
step_learning_model_stanc <- stanc(
  file = "stan-models/step_learning_model.stan",
  model_name = "step_learning_model")
step_learning_model <- stan_model(
  stanc_ret = step_learning_model_stanc)
save(step_learning_model,
  file = paste0("stan-models/compiled/", os,
    "/step_learning_model.Rds"),
  compress = "xz", compression_level = 9)

# symmetric logistic learning
symmetric_logistic_learning_model_stanc <- stanc(
  file = "stan-models/symmetric_logistic_learning_model.stan",
  model_name = "symmetric_logistic_learning_model")
symmetric_logistic_learning_model <- stan_model(
  stanc_ret = symmetric_logistic_learning_model_stanc)
save(symmetric_logistic_learning_model,
  file = paste0("stan-models/compiled/", os,
    "/symmetric_logistic_learning_model.Rds"),
  compress = "xz", compression_level = 9)

# asymmetric logistic learning
asymmetric_logistic_learning_model_stanc <- stanc(
  file = "stan-models/asymmetric_logistic_learning_model.stan",
  model_name = "asymmetric_logistic_learning_model")
asymmetric_logistic_learning_model <- stan_model(
  stanc_ret = asymmetric_logistic_learning_model_stanc)
save(asymmetric_logistic_learning_model,
  file = paste0("stan-models/compiled/", os,
    "/asymmetric_logistic_learning_model.Rds"),
  compress = "xz", compression_level = 9)
```




<div class="indent2">
## No Learning Model {#def-no}

```{r load-no-learn, eval=TRUE}
load(paste0("stan-models/compiled/", os, "/no_learning_model.Rds"))
```

Assuming no learning occurs, this model is designed to fit the initial adaptation to the experimental task. This is the most basic model that we will test, as it defines the functional form for $\hat{\mu}$ as only a decaying exponential curve with a few parameters for fitting the data. The model for a single subject is:

\begin{equation} \label{eqn:model-no}
  \begin{aligned}
    \hat{\mu_n} &= \log\big( V + E \cdot \exp(-A \cdot t) \big),\\[2ex]
    \hat{\mu_\ell} &= \log\big( (V + S) + E \cdot \exp(-A \cdot t) \big),\\[2ex]
  \end{aligned}
  \tag{M.1}
\end{equation}

where $t$ is the trial iteration, $t \in \{1, \dots, T\}$, and $T$ is the number of trials performed in the experiment by that individual. The parameters are explained in detail in the following list:

* $V \in [0, 2]$ represents the median pre-learning response time in seconds; technically, it is the vertical shift of all response times. Naturally, it should not be negative because response times cannot be negative in this experiment. Furthermore, the maximum value is 2 seconds because each letter was only visible in the window for 2 seconds before the next letter appeared and the response time was cut off.
* $E \in [0, \infty)$ is the overall scale of the exponential used to model the period of adaptation to the task. It must be non-negative because it must be monotonically decreasing as the subject gets used to the experiment.
* $A \in [0, \infty)$ is the adaptation rate to the task. It again must be non-negative so that the exponential term decays instead of grows.
* $S \in (-\infty, \infty)$ is the shift in response time of the learned letters \emph{prior to learning} relative to that of the non-learned letters. We expect this value to be close to zero since essentially controls for any mechanical differences in typing the different letters on the keyboard.
* $\sigma^2 \in [0, \infty)$ drives the variance in the response times more than $\mu'$, so we treat it as a proxy for the variance; naturally, it should be non-negative.

The prior distributions for the model parameters are:

\begin{equation} \label{eqn:model-no-prior}
  \begin{aligned}
    V &\sim \text{Gamma}(2.5, 2.5),\\[0.25ex]
    E &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    A &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    S &\sim \text{Normal}(0, 0.5),\\[0.25ex]
    \sigma &\sim \text{Gamma}(2, 10).
  \end{aligned}
  \nonumber
\end{equation}

You might notice that the Gamma prior on $V$ technically has support for any non-negative real number, yet the response times themselves are bounded above by 2 seconds. There aren't any useful priors that are defined on only $[0, 2]$, but we could use a prior bounded on $[0, 1]$ and scale it to support $[0, 2]$ (such as a Beta distribution). I opted for the Gamma prior so that the model would be more easily generalizable for other data, but the Stan model defines an upper bound of 2 seconds when fitting $V$ so any fitted value of $V$ can be at most $2$.

We provide a fairly sharp Normal prior on $S$ so that it remains close to zero as we do not expect any major difference in the mechanical differences in typing on the keyboard. The other priors are weakly informative distributions based on their domains; since the priors are relatively weak, we allow for the data to drive the fitted parameter values instead of our prior beliefs. The priors on $E$, $A$, and $\sigma$ are simple Gamma distributions with ample room for exploring the possible parameter values.




## Step Learning Model {#def-step}

```{r load-step-learn, eval=TRUE}
load(paste0("stan-models/compiled/", os, "/step_learning_model.Rds"))
```

This model builds on the previous decaying exponential by including a sudden drop in the median response time to allow for fitting the evidence of learning. We add two parameters: $D$, the scale of the drop in median response time, and $H$ the onset of learning. The sudden drop in response time takes the shape of a step function, which assumes that the learning occurs across the timespan of a single trial in the experiment. Whereas this model fits well to simple learning tasks where the learning occurs nearly instantaneously, it is rather inflexible for a more general learning process. The model for a single subject is:

\begin{equation} \label{eqn:models-step}
  \begin{aligned}
    \hat{\mu_n} &= \log\big( V + E \cdot \exp(-A \cdot t) \big),\\[2ex]
    \hat{\mu_\ell} &= \log\big( (V + S) + E \cdot \exp(-A \cdot t) - ({\bf 1}_{t \geq H} \cdot D) \big),\\[2ex]
  \end{aligned}
  \tag{M.2}
\end{equation}

where $t$ is the trial iteration, $t \in \{1, \dots, T\}$, $T$ is the number of trials performed in the experiment by that individual, and ${\bf 1}_{t \geq H}$ is the indicator function that resolves to $1$ if $t \geq H$ and $0$ if $t < H$. In the coding of the model, care is taken to isolate the parameters $S$ and $D$ so that they do not get confounded; more precisely, the parameter $D$ is only applicable after $H$ trials have been completed by the participant. All of the parameters are explained in detail in the following list:

* $V \in [0, 2]$ represents the median pre-learning response time in seconds; technically, it is the vertical shift of all response times. Naturally, it should not be negative because response times cannot be negative in this experiment. Furthermore, the maximum value is 2 seconds because each letter was only visible in the window for 2 seconds before the next letter appeared and the response time was cut off.
* $E \in [0, \infty)$ is the overall scale of the exponential used to model the period of adaptation to the task. It must be non-negative because it must be monotonically decreasing as the subject gets used to the experiment.
* $A \in [0, \infty)$ is the adaptation rate to the task. It again must be non-negative so that the exponential term decays instead of grows.
* $S \in (-\infty, \infty)$ is the shift in response time of the learned letters \emph{prior to learning} relative to that of the non-learned letters. We expect this value to be close to zero since essentially controls for any mechanical differences in typing the different letters on the keyboard.
* $D \in [0, 1]$ is the scale of the learning "drop." It is the scaled difference between the median pre-learning response time and the median post-learning response time for the learned letter in the pattern. A value near $0$ indicates a no decrease in response time (a post-learning response time equal to that of the pre-learning); a value near $1$ indicates a full decrease in response time (a post-learning response time of $0$ seconds).
* $H \in [0, T]$, is the onset of learning, where $T$ is the number of trials performed in the experiment by the individual. $H$ represents how many trials it takes for the subject to learn the existence of the pattern.
* $\sigma^2 \in [0, \infty)$ drives the variance in the response times more than $\mu'$, so we treat it as a proxy for the variance; naturally, it should be non-negative.

The prior distributions for the model parameters are:

\begin{equation} \label{eqn:model-step-prior}
  \begin{aligned}
    V &\sim \text{Gamma}(2.5, 2.5),\\[0.25ex]
    E &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    A &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    S &\sim \text{Normal}(0, 0.5),\\[0.25ex]
    D &\sim \text{Beta}(2.5, 2.5),\\[0.25ex]
    H &\sim \text{Cauchy}(\tfrac{T}{2}, 25),\\[0.25ex]
    \sigma &\sim \text{Gamma}(2, 10).
  \end{aligned}
  \nonumber
\end{equation}

In addition to the priors already covered in the previous section, we place a slightly informative Beta prior on $D$ in order to dissuade the model from fitting values too close to either $0$ or $1$. Whereas values close to $1$ are extremely unlikely, values close to $0$ can confound a consistent drop in the median response time with random noise, thus not identifying the underlying cognitive process.

Moreover, the prior on $H$ is also slightly informative as we discourage the model from fitting the onset of learning at the very beginning or the very end of the experiment. In cases such as these, we cannot distinguish legitimate evidence of learning behavior from simple adaptation to the task or random noise at the end of the experiment. While the Cauchy prior does technically support any real number (including negative numbers and zero), the Stan model bounds $H$ below by $0$ and above by $T$, the number of trials completed by the individual (so the upper bound is different for each participant). Ultimately, this Cauchy prior suggests that the learning occurs somewhere around halfway through the experiment, but it doesn't prevent the model from fitting fast or slow learners.




## Symmetric Logistic Learning Model {#def-sym}

```{r load-sym-log-learn, eval=TRUE}
load(paste0("stan-models/compiled/", os, "/symmetric_logistic_learning_model.Rds"))
```

This model adds another layer of complexity to the Step Learning Model because it allows the drop in the median response time to be gradual by including the additional parameter $L$, the learning rate. This model can approach a step function by allowing the learning rate to increase to $\infty$ (i.e., assuming that the learning happens instantaneously), yet it is still able to characterize gradual learning by fitting the learning rate to a value less than $1$. Again, the parameter $L$ does not affect the model until $H$ trials have been completed by the participant. The model for a single subject is:

\begin{equation} \label{eqn:models-sym}
  \begin{aligned}
    \hat{\mu_n} &= \log\big( V + E \cdot \exp(-A \cdot t) \big),\\[2ex]
    \hat{\mu_\ell} &= \log\bigg( (V + S) + E \cdot \exp(-A \cdot t) - \cdot \Big( 1 - \frac{D}{1+\exp(-L \cdot (t-H))} \Big) \bigg),\\[2ex]
  \end{aligned}
  \tag{M.3}
\end{equation}

where $t$ is the trial iteration, $t \in \{1, \dots, T\}$, and $T$ is the number of trials performed in the experiment by that individual. All of the parameters are explained in detail in the following list:

* $V \in [0, 2]$ represents the median pre-learning response time in seconds; technically, it is the vertical shift of all response times. Naturally, it should not be negative because response times cannot be negative in this experiment. Furthermore, the maximum value is 2 seconds because each letter was only visible in the window for 2 seconds before the next letter appeared and the response time was cut off.
* $E \in [0, \infty)$ is the overall scale of the exponential used to model the period of adaptation to the task. It must be non-negative because it must be monotonically decreasing as the subject gets used to the experiment.
* $A \in [0, \infty)$ is the adaptation rate to the task. It again must be non-negative so that the exponential term decays instead of grows.
* $S \in (-\infty, \infty)$ is the shift in response time of the learned letters \emph{prior to learning} relative to that of the non-learned letters. We expect this value to be close to zero since essentially controls for any mechanical differences in typing the different letters on the keyboard.
* $D \in [0, 1]$ is the scale of the learning "drop." It is the scaled difference between the median pre-learning response time and the median post-learning response time for the learned letter in the pattern. A value near $0$ indicates a no decrease in response time (a post-learning response time equal to that of the pre-learning); a value near $1$ indicates a full decrease in response time (a post-learning response time of $0$ seconds).
* $L \in [0, \infty)$ is the learning rate of the subject. A negative learning rate would cause the logistic part of the expression to increase rather than decrease. A learning rate approaching $\infty$ approximates a step function (i.e., learning occurs instantaneously).
* $H \in [0, T]$, is the onset of learning, where $T$ is the number of trials performed in the experiment by the individual. $H$ represents how many trials it takes for the subject to learn the existence of the pattern.
* $\sigma^2 \in [0, \infty)$ drives the variance in the response times more than $\mu'$, so we treat it as a proxy for the variance; naturally, it should be non-negative.

The prior distributions for the model parameters are:

\begin{equation} \label{eqn:model-sym-prior}
  \begin{aligned}
    V &\sim \text{Gamma}(2.5, 2.5),\\[0.25ex]
    E &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    A &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    S &\sim \text{Normal}(0, 0.5),\\[0.25ex]
    D &\sim \text{Beta}(2.5, 2.5),\\[0.25ex]
    L &\sim \text{Gamma}(1.5, 0.25),\\[0.25ex]
    H &\sim \text{Cauchy}(\tfrac{T}{2}, 25),\\[0.25ex]
    \sigma &\sim \text{Gamma}(2, 10).
  \end{aligned}
  \nonumber
\end{equation}

The prior distribution that we place on $L$ is weakly informative, and it allows for fitting relatively large values of $L$ (e.g., $L > 5$) that cause the model to closely mimic a step function. This prior also allows the model to fit a smaller value for $L$, characterizing a more gradual learning process.




## Asymmetric Logistic Learning Model {#def-asym}

```{r load-asym-log-learn, eval=TRUE}
load(paste0("stan-models/compiled/", os, "/asymmetric_logistic_learning_model.Rds"))
```

The final model is very similar to the previous Symmetric Logistic Learning Model, but it uses the generalized form of the logistic function to model the shape of the decrease in median response time. By adding three additional parameters ($C$, $Q$, and $\nu$), this model allows the drop in the median response time to be not only gradual but also asymmetric. Again, these new parameters do not affect the model until $H$ trials have been completed by the participant. The model for a single subject is:

\begin{equation} \label{eqn:models-asym}
  \begin{aligned}
    \hat{\mu_n} &= \log\big( V + E \cdot \exp(-A \cdot t) \big),\\[2ex]
    \hat{\mu_\ell} &= \log\bigg( (V + S) + E \cdot \exp(-A \cdot t) - \cdot \Big( 1 - \frac{D}{1+\exp(-L \cdot (t-H))} \Big) \bigg),\\[2ex]
  \end{aligned}
  \tag{M.3}
\end{equation}

where $t$ is the trial iteration, $t \in \{1, \dots, T\}$, and $T$ is the number of trials performed in the experiment by that individual. All of the parameters are explained in detail in the following list:

* $V \in [0, 2]$ represents the median pre-learning response time in seconds; technically, it is the vertical shift of all response times. Naturally, it should not be negative because response times cannot be negative in this experiment. Furthermore, the maximum value is 2 seconds because each letter was only visible in the window for 2 seconds before the next letter appeared and the response time was cut off.
* $E \in [0, \infty)$ is the overall scale of the exponential used to model the period of adaptation to the task. It must be non-negative because it must be monotonically decreasing as the subject gets used to the experiment.
* $A \in [0, \infty)$ is the adaptation rate to the task. It again must be non-negative so that the exponential term decays instead of grows.
* $S \in (-\infty, \infty)$ is the shift in response time of the learned letters \emph{prior to learning} relative to that of the non-learned letters. We expect this value to be close to zero since essentially controls for any mechanical differences in typing the different letters on the keyboard.
* $D \in [0, 1]$ is the scale of the consistent decrease in median response time upon learning the existence of the pattern. It is the scaled difference between the median pre-learning response time and the median post-learning response time for the learned letter in the pattern. A value near $0$ indicates a no decrease in response time (a post-learning response time equal to that of the pre-learning); a value near $1$ indicates a full decrease in response time (a post-learning response time of $0$ seconds).
* $L \in [0, \infty)$ is the learning rate of the subject. A negative learning rate would cause the logistic part of the expression to increase rather than decrease. A learning rate approaching $\infty$ approximates a step function (i.e., learning occurs instantaneously).
* $H \in [0, T]$, is the onset of learning, where $T$ is the number of trials performed in the experiment by the individual. $H$ represents how many trials it takes for the subject to learn the existence of the pattern.
* $C \in [0, \infty)$ inversely affects the scale of the consistent decrease in median response time upon learning the existence of the pattern.
* $Q \in [0, \infty)$ affects the onset of learning.
* $\nu \in [0, \infty)$ affects near which asymptote the maximum learning occurs (i.e., if learning occurs faster earlier or later in the learning process); this affect is only visible if the learning rate, $L$, is small. In addition, it also affects the scale of the consistent decrease in median response time upon learning the existence of the pattern.
* $\sigma^2 \in [0, \infty)$ drives the variance in the response times more than $\mu'$, so we treat it as a proxy for the variance; naturally, it should be non-negative.

The prior distributions for the model parameters are:

\begin{equation} \label{eqn:model-asym-prior}
  \begin{aligned}
    V &\sim \text{Gamma}(2.5, 2.5),\\[0.25ex]
    E &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    A &\sim \text{Gamma}(2.5, 10),\\[0.25ex]
    S &\sim \text{Normal}(0, 0.5),\\[0.25ex]
    D &\sim \text{Beta}(2.5, 2.5),\\[0.25ex]
    L &\sim \text{Gamma}(1.5, 0.25),\\[0.25ex]
    H &\sim \text{Cauchy}(\tfrac{T}{2}, 25),\\[0.25ex]
    C &\sim \text{Log-Normal}(1, 1),\\[0.25ex]
    Q &\sim \text{Log-Normal}(1, 1),\\[0.25ex]
    \nu &\sim \text{Log-Normal}(1, 1),\\[0.25ex]
    \sigma &\sim \text{Gamma}(2, 10).
  \end{aligned}
  \nonumber
\end{equation}

All three of the prior distributions that we place on the newly added parameters ($C$, $Q$, and $\nu$) are slightly informative Log-Normal distributions centered at $1$. We chose the Log-Normal distribution because it only has support for non-negative real numbers, precisely the domain of the three newly added parameters. Moreover, these distributions are centered on $1$ but still allow the model to fit more extreme values. Note that if each of these parameters is $1$ ($C = Q = \nu = 1$), then we return to the simpler symmetric logistic learning model in the previous subsection.
</div>





# Model Recovery {#rec}

This section will contain some basic model recovery experiments in order to determine each model's flexibility when fitting data that exhibit distinct patterns. In each of the following subsections, we will choose the response time data from one subject in the experiment that most clearly exhibits a specific behavioral pattern. Then we will fit this selected data to each of the models from the previous section in order to demonstrate each model's ability to capture commonly observed behaviors. We will use data for the response times of both the first letter in the pattern and the third letter in the pattern; if there is evidence of learning the pattern, we expect to see a decrease in the median response time for the third letter and not for the first letter. First we read the data from file so that we can later select individual participants' response time data.

```{r load-data, eval=TRUE}
load(file = "experiment-1/kendal-model/exp1.Rds")
```

<div class="indent2">
## No Evidence of Learning {#rec-nolearn}

The first behavioral pattern that we will test is from a participant that exhibits no evidence of learning the lexicographical pattern in the sequence of letters. This data will contain the adaptation to the experimental task, as shown by a decrease in response time across the first several trials of the experiment. Following that, however, will be no further changes to the median response time. Overall, this data should heavily favor the No Learning Model. We will use `subject_id = 171` for this subsection; another consideration would be subject 130.

<div class="indent3">
### No Learning Model {#rec-nolearn-no}
### Step Learning Model {#rec-nolearn-step}
### Symmetric Logistic Learning Model {#rec-nolearn-sym}
### Asymmetric Logistic Learning Model {#rec-nolearn-asym}
</div>




## Strong Evidence of Learning {#rec-learn}

Second, we will use response time data from a participant that exhibits strong evidence of learning the lexicographical pattern in the sequence of letters. This data will not only contain the adaptation to the experimental task, but it will also contain a significant decrease in the median response time of the learned letter approximately two-thirds of the way through the experiment. We expect the No Learning Model to struggle with this data while the other three models should be able to capture the behavior quite well. Here we will use `subject_id = 74`, although subject 204 is another option.

<div class="indent3">
### No Learning Model {#rec-learn-no}
### Step Learning Model {#rec-learn-step}
### Symmetric Logistic Learning Model {#rec-learn-sym}
### Asymmetric Logistic Learning Model {#rec-learn-asym}
</div>




## Consistent but Small Decrease in Response Times {#rec-small}

In this subsection, we will test the models with data from a participant that shows a small, but consistent, decrease in their response times. This behavior suggests that the subject may have learned the lexicographical pattern, but their response times did not decrease as significantly. Data of this shape can be tricky as the small decrease in response time presents a difficult distinction between whether the No Learning Model should fit better than the three learning models. For this data, we will use 'subject_id = 142', and subjects 237 and possibly 143 are good as well.

<div class="indent3">
### No Learning Model {#rec-small-no}
### Step Learning Model {#rec-small-step}
### Symmetric Logistic Learning Model {#rec-small-sym}
### Asymmetric Logistic Learning Model {#rec-small-asym}
</div>




## Sporadic Return to Baseline Response Time {#rec-ret}
28 152 235
This subsection will focus on when the response times of the learned third letter occasionally revert back to the baseline prior to learning. This sporadic behavior could show evidence of another possible consequence of a loss of focus during the experiment, but it could also be something more innocent such as temporarily forgetting the pattern. Since the models do not have a built in mechanism to capture this seemingly stochastic behavior, they will likely fit somewhere in between , possibly approaching a similar fit to the small but consistent decrease in response time from the previous section. The participant that most clearly demonstrates this behavior is `subject_id = 28`, but subjects 152 and 235 are also options.


<div class="indent3">
### No Learning Model {#rec-ret-no}
### Step Learning Model {#rec-ret-step}
### Symmetric Logistic Learning Model {#rec-ret-sym}
### Asymmetric Logistic Learning Model {#rec-ret-asym}
</div>




## Increased Response Times at the End of the Experiment {#rec-inc}

An inevitable feature of lengthy and repetitive laboratory experiments is that the participant eventually gets bored of completing the same task over and over again. This loss of focus can lead to lack of thought and care in the participant's responses; but in particular to this experiment, we observe increased response times relative to their normal baseline. A significant increase in the median response time toward the end of the experiment can hinder the models' abilities to accurately capture the true learning (or lack of learning) that occurs throughout the experiment. We expect that the models will fit slightly higher post-learning median response times to accommodate for this late increase. We will use `subject_id = 149` for this testing, but alternatives are subjects 86 and 50.

<div class="indent3">
### No Learning Model {#rec-inc-no}
### Step Learning Model {#rec-inc-step}
### Symmetric Logistic Learning Model {#rec-inc-sym}
### Asymmetric Logistic Learning Model {#rec-inc-asym}
</div>




## Other Considerations {#rec-other}
both sporadic return and boredom: 187

boredom and sporadic return at the end (perhaps because of boredom) (with learning): 7
</div>






# Formal Model Comparison (just a sample for now) {#comp}

<div class="indent2">
## Posterior Predictive Fits {#comp-ppf}

## Bayes Factor {#comp-bayes}

## Loo {#comp-loo}

## Stacking Weights {#comp-weights}
</div>





# Sample Fits {#fits}
<hr class="sec1">

These are sample fits using the model. We'll show simulated data for parameter recovery and real data collected from the experiment.




## Fitting Simulated Data {#fits-sim}

Here we're basically just checking to make sure that the model recovers the parameters from the simulated data.

First we read the model into R using the `stanc` function; then we compile the model using the `stan_model` function. You can use a shortcut function to accomplish all of this and more in one function call, but I prefer to do it sequentially because I find it easier to catch bugs and other errors. Disregard the warnings about $H$- it's just because of how Stan makes you define the variable upper bounds.

```{r stan-model, eval=TRUE}
library("rstan")
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

exp_log_model_stanc <- stanc(file = "../exp_logistic_model.stan",
                             model_name = "exponential_logistic_model")
exp_log_model <- stan_model(stanc_ret = exp_log_model_stanc)
```

```{r stan-model-internal, eval=FALSE, include=FALSE}
save(exp_log_model, file = "../exp_logistic_model.Rds",
     compress = "xz", compression_level = 9)
load("../exp_logistic_model.Rds")
```

We'll define the functions that we'll use to generate some simulated data.

```{r exp-log-funcs, eval=TRUE}
exp_log <- function(trial, V, E, A, D, L, H) {
  return( (V + E*exp(-A*trial)) * (1 - D/(1 + exp(-L*(trial-H)))) )
}
sim_exp_log <- function(trial, V, E, A, D, L, H, var) {
  sim <- exp_log(trial = trial, E = E, A = A, V = V, D = D, L = L, H = H) +
         rnorm(length(trial), 0, var)
  sim[sim < 0] <- 0
  return(sim)
}
```

Let's pick some parameter values and generate some simulated data. We're making three participants: the first one learns the pattern after 50 trials, the second one does not learn the pattern, and the third one shows a very small decrease in response time after 30 trials. The last participant is there just to see how the model handles a tricky situation. I have not included the observed behavior of demonstrating evidence of learning the pattern yet sporadically producing response times that are consistent with the pre-learning median.

```{r exp-log-gen, eval=TRUE}
K <- 3 # number of individuals
N <- 100 # number of trials for each individual (same for all)
NTI <- rep(N, K) # so Stan can handle multiple individuals
NK <- N*K
Y0 <- rep(NA, NK) # vector of non-learned response times
Y1 <- rep(NA, NK) # vector of learned response times
V <- c(1.1, 1.5, .75)
E <- c(.25, .25, .25)
A <- c(.25, .25, .25)
P <- c(1, 0, 1)
D <- c(.4, 0, .1)
L <- c(.4, .4, .4)
H <- c(50, 60, 30)
st <- 0
for (i in 1:K) {
  x <- seq_len(NTI[i])
  Y0[(st+1):(st+NTI[i])] <- sim_exp_log(x, V=V[i], E=E[i], A=A[i], D=0, L=L[i], H=H[i], var=0.15) # generate non-learned response times
  Y1[(st+1):(st+NTI[i])] <- sim_exp_log(x, V=V[i], E=E[i], A=A[i], D=D[i], L=L[i], H=H[i], var=0.15) # generate learned response times
  st = st + NTI[i]
}
```

Now we'll actually fit our generated data with our model. This a quick and dirty fit. It's only running one chain, not a ton of iterations, and not incredibly strict controls with regard to how closely it attempts to fit the data. It will most certainly throw a lot of warnings, but in my experience these quick fits are still quite good and take under 30 seconds per individual. I'm just going to use the summary of the fit so we can easily see how well the parameters are recovered. Note that if you want to modify the R code and only fit one individual, it's important to input `NTI` with `as.array()` because Stan can't handle vectors of length one so it's got to be an array instead. If you input more than one individual, you don't need the `as.array()`, but it doesn't hurt. Once the fit has run, we'll print some of the summary statistics of the fit so we can see the means and standard deviations of the parameter estimates.

```{r sim-fit, eval=TRUE}
sim_fit <- summary(sampling(
  exp_log_model,
  data = list('K'=K, 'NTI'=as.array(NTI), 'NK'=NK, 'Y0'=Y0, 'Y1'=Y1),
  refresh = FALSE, chains = 1, iter = 500, seed = 2,
  control = list(adapt_delta = 0.9, max_treedepth = 10)
))$summary

sim_fit[, c(1, 3)]
```

Printing the summary tells us about the parameter estimates, but I think a visualization could be more useful. We'll plot the mean estimate for each parameter, surrounded by the 2.5\% and 97.5\% confidence intervals from the fitting. On top of that we'll plot the true value that was input to the simulation.

```{r plot-sim-fit-pars, eval=TRUE}
plot_sim_fit_pars <- function(fits, K, V, E, A, P, D, L, H) {
  pars <- data.frame(id = seq_len(K),
                     V = V,
                     E = E,
                     A = A,
                     P = P,
                     D = D,
                     L = L,
                     H = H)
  idx <- c(1:6, 9)
  for (i in seq_len(length(idx))) {
    df <- as.data.frame(fits[seq_len(K) + (idx[i]-1)*K, c(1, 4, 8)])
    colnames(df) <- c("mean", "qlow", "qupp")
    df$id <- seq_len(K)
    df$dumm = c(1, rep(2, K-1))
    par_name <- substring(rownames(df)[1], 1, 1)

    print(ggplot(df) +
      geom_pointrange(alpha = 0.6, shape = 16, color = "#5fc6fa",
                      aes(x = id, y = mean, ymin = qlow, ymax = qupp,
                          size = factor(dumm, levels = c(1, 2)))) +
      geom_point(data = pars, alpha = 0.8, color = "#04547c",
                 shape = 4, size = 6, stroke = 2,
                 aes_string(x = "id", y = par_name)) +
      scale_size_manual(values = c(1.51, 1.5),
                        labels = c("True Value", "Est. Mean"),
                        name = NULL) +
      guides(size = guide_legend(override.aes = list(size = c(1.5, 1.5),
                                                     stroke = c(2, 1),
                                                     shape = c(4, 16),
                                                     lty = c(0, 0),
                                                     color = c("#04547c",
                                                               "#5fc6fa")))) +

      scale_x_discrete(limits = as.character(seq_len(K))) +
      labs(title = paste0("Simulated Parameter Estimates for ", par_name),
           x = "ID of Individual", y = "Estimated Parameter Value",
           subtitle = "Bars represent the 2.5% and 97.5% CI") +
      theme_bw() +
      theme(panel.border = element_blank(),
            plot.title = element_text(size = 20),
            plot.subtitle = element_text(size = 16,
                                         margin = margin(0, 5, 15, 5, "pt")),
            axis.text.x = element_text(size = 14),
            axis.text.y = element_text(size = 14),
            axis.title.x = element_text(size = 16,
                                        margin = margin(10, 0, 0, 0, "pt")),
            axis.title.y = element_text(size = 16,
                                        margin = margin(0, 10, 0, 0, "pt")),
            legend.position = c(1, 1),
            legend.justification = c(1, 0),
            legend.box = "horizontal",
            legend.direction = "vertical",
            legend.background = element_rect(fill = "transparent"))
    )
  }
}

plot_sim_fit_pars(sim_fit, K, V, E, A, P, D, L, H)
```

As we suspected, some parameters are more difficult to fit than others. The parameters $E$ and $A$ are slightly loose, and I attribute this to a lack of data since the acclimation to the task should be very short. Still, the parameter estimates don't seem wildly off. At first glance, it appears that the parameters $D$, $L$, and $H$ have poor fits for individuals 2 and 3. This lack of fitting is expected, though, because these parameters have no meaning when $P$ is close to zero as there is no learning to fit. Individual 3 does pose problems for the model, and we'll see some more evidence of this in the next plot.

Now we'll plot the fits over the generated data to give a visual as to how well the fitted parameters match the data. We'll write a function to do the plotting so all of the temporary bits are nicely contained.

```{r plot-sim-fits, eval=TRUE}
plot_sim_fits <- function(fits, K, NTI,
                          V = NA, E = NA, A = NA, D = NA,
                          L = NA, H = NA, Y0 = NA, Y1 = NA,
                          savepath = NULL) {
  st <- 0
  for (i in 1:K) {
    n <- NTI[i]
    Vi <- fits[i+0*K]
    Ei <- fits[i+1*K]
    Ai <- fits[i+2*K]
    Pi <- fits[i+3*K]
    Di <- fits[i+4*K]
    Li <- fits[i+5*K]
    Hi <- fits[i+8*K]
    sigma_2i <- fits[i+7*K]
    sn <- seq_len(n)

    df <- data.frame(x = rep(sn, 2),
                     data = c(Y0[(st+1):(st+n)], Y1[(st+1):(st+n)]),
                     truth = c(exp_log(sn, V = V[i], E = E[i], A = A[i],
                                       D = 0, L = L[i], H = H[i]),
                               exp_log(sn, V = V[i], E = E[i], A = A[i],
                                       D = D[i], L = L[i], H = H[i])),
                     fit = c(exp_log(sn, V = Vi, E = Ei, A = Ai,
                                     D = 0, L = Li, H = Hi),
                             exp_log(sn, V = Vi, E = Ei, A = Ai,
                                     D = Pi * Di, L = Li, H = Hi)),
                     data_label = c(rep(1,n), rep(2,n)),
                     fit_label = c(rep(3,n), rep(4,n)))
    df$fit_min <- df$fit - df$fit*sqrt(sigma_2i*(sigma_2i - 1))
    df$fit_max <- df$fit + df$fit*sqrt(sigma_2i*(sigma_2i - 1))
    df$dumm <- c(rep(1, n), rep(2, n))

    # factor levels key (since R makes it alphabetical)
      # 1: Non-Learned Data/Truth
      # 2: Learned Data/Truth
      # 3: Non-Learned Fit
      # 4: Learned Fit

    print(ggplot(df) +
      geom_ribbon(linetype="blank", alpha=0.25,
                  aes(x = x, ymin = fit_min, ymax = fit_max,
                      fill = factor(fit_label, levels = c(3, 4)))) +
      geom_point(alpha = 0.75, shape = 16,
                 aes(x = x, y = data,
                     color = factor(data_label, levels = c(1, 2)),
                     size = factor(dumm, levels = c(1, 2)))) +
      geom_line(size = 1.25, alpha = 0.8,
                aes(x = x, y = truth,
                    color = factor(data_label, levels = c(1, 2)))) +
      geom_line(size = 1.25, alpha = 0.6,
                aes(x = x, y = fit,
                    color = factor(fit_label, levels = c(3, 4)))) +
      scale_color_manual(values = c("#999999", "#000000",
                                    "#98c1ff", "#ff4f4f"),
                         labels = c("Non-Learned Truth", "Learned Truth",
                                    "Non-Learned Fit", "Learned Fit"),
                         name = NULL) +
      scale_fill_manual(values = c("#98c1ff", "#ff4f4f"),
                        guide = FALSE) +
      scale_size_manual(values = c(1.8, 1.81),
                        labels = c("Non-Learned Data", "Learned Data"),
                        name = NULL) +
      guides(color = guide_legend(order = 2,
                                  override.aes = list(shape = NA,
                                                      size = 1.25,
                                                      lty = rep(1, 4))),
             size = guide_legend(order = 1,
                                 override.aes = list(size = 2,
                                                     shape = c(16, 16),
                                                     color = c("#999999",
                                                               "#000000")))) +
      labs(title = "Simulated Data Fit",
           x = "Trial Number", y = "Response Time (sec)",
           subtitle = paste0("\u00B1 1 standard deviation\n",
                             "Subject ", i, "\n",
                             "P = ", round(Pi, 3))) +
      theme_bw() +
      theme(panel.border = element_blank(),
            plot.title = element_text(size = 20),
            plot.subtitle = element_text(size = 16,
                                         margin = margin(0, 5, 15, 5, "pt")),
            axis.text.x = element_text(size = 14),
            axis.text.y = element_text(size = 14),
            axis.title.x = element_text(size = 16,
                                        margin = margin(10, 0, 0, 0, "pt")),
            axis.title.y = element_text(size = 16,
                                        margin = margin(0, 10, 0, 0, "pt")),
            legend.position = c(1, 1),
            legend.justification = c(1, 0),
            legend.box = "horizontal",
            legend.direction = "vertical",
            legend.background = element_rect(fill = "transparent"))
    )

    if (!is.null(savepath)) {
      ggsave(paste0(savepath, "sim_fit_", i, ".png"))
    }

    st = st + n
  }
}

plot_sim_fits(sim_fit, K, NTI, V, E, A, D, L, H, Y0, Y1)
```

We can see that everything looks pretty good except for the last fake participant who showed just a very small decrease in median response time. The model finds it difficult to determine where the decrease occurs, and it also confuses $P$ and $D$ a little bit. Just by looking at the plot, I wouldn't say that the individual had learned the pattern, and I think such a small decrease in response time is a potential problem area for this model. However, I don't anticipate this trend appearing in the data very frequently, and I also can't see a better way of handling such data besides saying "the individual may have showed some signs of learning the pattern, but nothing really." Perhaps in practice this problematic fit could indicate that the individual exhibits atypical learning patters, and maybe that could be useful as a flag to engage with them on a more personal level?





## Fitting Real Experimental Data {#fits-real}

```{r examine-data-internal, eval=FALSE, include=FALSE}
# load data, will be in the variable 'exp1'
load(file = "../exp1.Rds")
sub_ids <- unique(exp1$subject_id)

par(mfrow=c(length(sub_ids),1))
for (i in 1:length(sub_ids)) {
  s0 <- subset(exp1, subject_id == sub_ids[i] & is_predictable==0)$rt/1000
  s1 <- subset(exp1, subject_id == sub_ids[i] & is_predictable==1)$rt/1000
  n <- min(length(s0), length(s1))
  s0 <- s0[seq_len(n)]
  s1 <- s1[seq_len(n)]
  x <- seq_len(n)
  plot(x, s1, pch=20, col='black', ylab = paste("Subject", i, sep=" "))
  points(x, s0, pch=20, col='gray50')
}
```

Now we'll use the model to fit data from the actual experiment. We'll take three individuals who appear to have differing levels of learning: the first will show clear evidence of learning, the second will show no evidence of learning, and the third will show some evidence of learning but will sporadically revert back to their pre-learning median response time. First we'll load the data from an RDS file because it can be compressed smaller than a CSV, and then we'll prep everything that we need to run the fitting. Note that as a precaution, we remove all the individuals with fewer than 50 trials (all of the selected individuals have more than 50 trials, but I wanted to keep the code generalizable).

```{r data-prep, eval=TRUE}
# load data, will be in the variable 'exp1'
load(file = "../exp1.Rds")

# use these 3 individuals
sub_ids <- unique(exp1$subject_id)[c(78, 7, 10)]

# prep for fitting
K <- length(sub_ids)
gs_idx <- vector()
NTI <- rep(0, K)
Y0 <- vector()
Y1 <- vector()

for (i in seq_len(K)) {
  temp <- exp1[exp1$subject_id == sub_ids[i], ]
  temp0 <- temp[temp$is_predictable == 0, ]$rt/1000
  temp1 <- temp[temp$is_predictable == 1, ]$rt/1000
  mm <- min(length(temp0), length(temp1))
  if (mm >= 50) {
    NTI[i] <- mm
    Y0 <- c(Y0, temp0[seq_len(mm)])
    Y1 <- c(Y1, temp1[seq_len(mm)])
    gs_idx <- c(gs_idx, i)
  }
}

good_sub_ids <- sub_ids[gs_idx]
NTI <- NTI[NTI > 0]
NK <- sum(NTI)
K <- length(good_sub_ids)
```

Now that everything is ready to go, let's fit the experimental data. Again, this is a quick and dirty fit with all the precautions from the earlier fit.

```{r data-fit, eval=TRUE}
real_fit <- summary(sampling(
  exp_log_model,
  data = list('K' = K, 'NTI' = as.array(NTI), 'NK' = NK, 'Y0' = Y0, 'Y1' = Y1),
  refresh = FALSE, chains = 1, iter = 500, seed = 2,
  control = list(adapt_delta = 0.9, max_treedepth = 10)
))$summary

real_fit[, c(1, 3)]
```

Since we don't actually have the true value of the model parameters, we'll just plot how the fit compares to the data. Again, we'll wrap it all in a function and use that.

```{r plot-real-fits, eval=TRUE}
plot_real_fits <- function(fits, K, NTI, Y0 = NA, Y1 = NA, savepath = NULL) {
  st <- 0
  for (i in 1:K) {
    n <- NTI[i]
    Vi <- fits[i+0*K]
    Ei <- fits[i+1*K]
    Ai <- fits[i+2*K]
    Pi <- fits[i+3*K]
    Di <- fits[i+4*K]
    Li <- fits[i+5*K]
    Hi <- fits[i+8*K]
    sigma_2i <- fits[i+7*K]
    sn <- seq_len(n)

    df <- data.frame(x = rep(sn, 2),
                     data = c(Y0[(st+1):(st+n)], Y1[(st+1):(st+n)]),
                     fit = c(exp_log(sn, V = Vi, E = Ei, A = Ai,
                                     D = 0, L = Li, H = Hi),
                             exp_log(sn, V = Vi, E = Ei, A = Ai,
                                     D = Pi * Di, L = Li, H = Hi)),
                     data_label = c(rep(1,n), rep(2,n)),
                     fit_label = c(rep(3,n), rep(4,n)))
    df$fit_min <- df$fit - df$fit*sqrt(sigma_2i*(sigma_2i - 1))
    df$fit_max <- df$fit + df$fit*sqrt(sigma_2i*(sigma_2i - 1))

    # factor levels key (since R makes it alphabetical)
      # 1: Non-Learned Data/Truth
      # 2: Learned Data/Truth
      # 3: Non-Learned Fit
      # 4: Learned Fit

    print(ggplot(df) +
      geom_ribbon(linetype="blank", alpha=0.25,
                  aes(x = x, ymin = fit_min, ymax = fit_max,
                      fill = factor(fit_label, levels = c(3, 4)))) +
      geom_point(alpha = 0.75, shape = 16, size = 1.8,
                 aes(x = x, y = data,
                     color = factor(data_label, levels = c(1, 2)))) +
      geom_line(size = 1.25, alpha = 0.6,
                aes(x = x, y = fit,
                    color = factor(fit_label, levels = c(3, 4)))) +
      scale_color_manual(values = c("#999999", "#000000",
                                    "#98c1ff", "#ff4f4f"),
                         labels = c("Non-Learned Data", "Learned Data",
                                    "Non-Learned Fit", "Learned Fit"),
                         name = NULL) +
      scale_fill_manual(values = c("#98c1ff", "#ff4f4f"),
                        guide = FALSE) +
      guides(color = guide_legend(override.aes = list(size = c(2, 2, 1.25, 1.25),
                                                      shape = c(16, 16, NA, NA),
                                                      lty = c(NA, NA, 1, 1)))) +
      labs(title = "Experimental Data Fit",
           x = "Trial Number", y = "Response Time (sec)",
           subtitle = paste0("\u00B1 1 standard deviation\n",
                             "Subject ", i, "\n",
                             "P = ", round(Pi, 3))) +
      theme_bw() +
      theme(panel.border = element_blank(),
            plot.title = element_text(size = 20),
            plot.subtitle = element_text(size = 16,
                                         margin = margin(0, 5, 15, 5, "pt")),
            axis.text.x = element_text(size = 14),
            axis.text.y = element_text(size = 14),
            axis.title.x = element_text(size = 16,
                                        margin = margin(10, 0, 0, 0, "pt")),
            axis.title.y = element_text(size = 16,
                                        margin = margin(0, 10, 0, 0, "pt")),
            legend.position = c(1, 1),
            legend.justification = c(1, 0),
            legend.box = "horizontal",
            legend.direction = "vertical",
            legend.background = element_rect(fill = "transparent"))
    )

    if (!is.null(savepath)) {
      ggsave(paste0(savepath, "real_fit_", i, ".png"))
    }

    st = st + n
  }
}

plot_real_fits(real_fit, K, NTI, Y0, Y1)
```

We see similar results compared to fitting the simulated data. The model identifies both the individual who showed evidence of learning the pattern, and the individual who showed no evidence of learning the pattern. The trouble again arises when the data show weird inconsistencies in response times. The third individual showed maybe some signs of learning the pattern fairly early on in the experiment, but the variance seems to increase upon adjusting to the task. The model struggles to fit this, and maybe we can treat this as an indication of a poorly performing participant who needs closer examination? I think the only way to fit a model to this type of behavior is to incorporate some kind of stochasticity into the model to allow for sporadic reversions back to the pre-learning median response time. The tricky thing is that we don't want to overfit the data and end up fitting a bunch of noise because the participants were distracted while completing the task. We also need to be careful about the model potentially confusing regular variance in the response times for this newfangled stochasticity. I suppose we could change the $\hat{\sigma^2}$ parameter to change upon learning, and allow that to increase so that the variance increases. I don't know if trying to fit bored people is particularly fruitful though, and if it's just the occassional response time that reverts back to the pre-learning median then I think the stochastic part should handle that well enough.







</div>





# References
